To be FAIR: Theory Specification Needs an Update

### 2025-06-07 manuscript.tex
update 2025-06-07 12:47:56.764452

  man, noextraspace,floatsintext]{apa6}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\makeatother
\keywords{fairtheory, meta science, theory formation, open science\newline\indent Word count: 9245}
\usepackage{lineno}

@@

This is a preprint paper, generated from Git Commit \# 7d4ab9cc. The first and senior author positions were fixed, and remaining author positions randomized using \texttt{set.seed(6999982);\ sample(c("Aaron",\ "Felix",\ "Noah",\ "Maximilian"))}.

The authors made the following contributions. Caspar J. Van Lissa: Conceptualization, Formal Analysis, Funding acquisition, Methodology, Project administration, Software, Supervision, Writing -- original draft, Writing -- review \& editing; Aaron Peikert: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Maximilian S. Ernst: Writing -- review \& editing, Investigation; Noah N.N. van Dongen: Writing -- review \& editing; Felix D. Schönbrodt: Conceptualization, Writing -- review \& editing; Andreas M. Brandmaier: Formal Analysis, Writing -- original draft, Writing -- review \& editing.

@@

\abstract{%
Innovations in open science and meta-science have focused on rigorous \emph{theory testing}, yet methods for specifying, sharing, and iteratively improving theories remain underdeveloped. To address these limitations, we introduce \emph{FAIR theory}: A standard for specifying theories as Findable, Accessible, Interoperable, and Reusable information artifacts. FAIR theories are Findable in well-established archives, Accessible in terms of availability and their ability to be understood, Interoperable for specific purposes, such as selecting control variables, and Reusable so that they can be iteratively improved through collaborative efforts. This paper adapts the FAIR principles for theory, reflects on the FAIRness of contemporary psychological practices, introduces a workflow for FAIRifying theory, and discusses FAIR theories' potential impact in terms of reducing research waste, enabling meta-research on theories' structure and development, and incorporating theory into reproducible research workflows -- from hypothesis generation to simulation studies. FAIR theory constitutes framework for specifying and refining theories, bridging a critical gap in open scholarly practices and supporting the renewed interest in theory development in psychology and beyond. FAIR theory builds on existing open science principles and infrastructures to provide a structured, cumulative framework for theory development, potentially increasing efficiency and potentially accelerating the pace of cumulative knowledge acquisition.
}

@@

The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable (M. D. Wilkinson et al., 2016) for both humans and computers\footnote{\phantomsection\label{testone}{As the colloquial use of these terms differs in important ways from their definition according to the FAIR principles, we capitalize these terms when we refer to a specific FAIR principle.}}.
Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software (Lamprecht et al., 2019) and computational workflows (S. R. Wilkinson et al., 2024).
This paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
To this end, we introduce ``FAIR theory'':
a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
compliant with the FAIR principles.
\phantomsection\label{test2}{FAIR theory has the potential to improve the efficiency of scholarly communication and}
accelerate cumulative knowledge acquisition.
We focus on applications in psychology, but the principles are relevant across the social sciences and beyond.

@@

The so-called ``replication crisis'' has prompted extensive scientific reforms (Lavelle, 2021; Scheel, 2022).
Concern that the abundance of non-replicable findings
was caused by undisclosed flexibility in analyses led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
These various practices ensure transparent and repeated testing of hypotheses by committing to an analysis plan in advance.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence (Scheel, Schijen, \& Lakens, 2021).
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:

Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so ambiguous that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.
A good example of this is found in ``self-determination theory'' (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures (Deci, 1971).
Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs---autonomy, competence, and relatedness---all still categorized as ``intrinsic'' (Ryan \& Deci, 2000).
The difference between these definitions becomes clear when considering the act of changing an infant's dirty diaper.
Under the original definition, one might hypothesize that caregivers are not intrinsically motivated to change diapers, as this is hardly a joyous experience.
Under the expanded definition, one might hypothesize the opposite, as the act may fulfill the need for relatedness.
The expanded definition thus enables SDT to absorb potentially falsifying evidence.

Scholars have raised concerns about the state of theory in psychology for nearly 50 years (Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
One main concern is that theories lack \emph{formalization} (Szollosi \& Donkin, 2021).
When theories are ambiguous and hence require either subjective interpretation or additional background knowledge,
it becomes difficult do derive precise predictions,
and therefore hard to falsify the theory.
A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
Despite these concerns, scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical processes of theory construction and improvement have been largely overlooked.
The present paper addresses this knowledge gap by applying, for the first time,
open science principles to psychological theory.
We apply introduce \emph{FAIR theory} to
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

In a progressive research program (Lakatos, 1971),
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature:
According to Kühberger, Fritz, and Scherndl (2014), 89.6\% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
Only 15\% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis (McPhetres et al., 2021).
The remaining 85\% of deductive studies lacked an explicit connection between theory and hypothesis.
Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
in which case it would be straightforward and important to make these explicit (Fried, 2020; Norouzi, Kleinberg, Vermunt, \& Van Lissa, 2024).
Or, perhaps some hypotheses are reported as part of entrenched research practices (Gigerenzer, Krauss, \& Vitouch, 2004),
but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected (Van Lissa et al., 2020).
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
Collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings (van Rooij \& Baggio, 2021).

Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely connected to, let alone derived from, theory.
Moreover, test results do not routinely contribute to the improvement or rejection of theories.
The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology ``lack the cumulative character of scientific knowledge. They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest'' (Meehl, 1978).
The present paper seeks to make theory more tangible and practical by instantiating it as a digital ``object'' that scholars can access, reuse, and update in their daily workflows.

@@

Definitions of theory are abundant, and are the subject of extensive scholarly debate.
Given that a pluriformity of definitions are consistent with FAIR theory principles,
our suggested approach is not limited to any one particular definition.
Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic (Winther, 2021).
The syntactic view describes theories as ``sets of sentences in a given logical domain language'' (Winther, 2021, ch.~2),
acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
We recognize the syntactic view in Meehl's (1990) hierarchy of ever-more specific ``statements'' a theory might contain:
statements about the types of entities postulated (i.e., ontology),
statements about causal connections between those entities,
statements about the functional form of those connections,
and statements about their specific numerical values (cf. Frankenhuis, Panchanathan, \& Smaldino, 2023; Guest, 2024).
The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
It shifts the focus from theories as collections of sentences to mathematical models.
The term ``model'' is not uniquely defined within the literature;
it has been described as a ``specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory'' (Fried, 2020).
This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
The pragmatic view holds that there might not be one structure or definition of scientific theories, but instead, definitions differ across scientific domains.
It also argues that nonformal aspects (e.g.~commonly used analogies) and practices (e.g.~experimental designs) can be an important part of scientific theories.

@@

We propose FAIR theory as an instantiation of modular publishing (Kircz, 1998).
In most fields, the primary unit of scientific communication is the academic paper.
A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
Modular publishing is the practice of making each of these resources available as independent citable \emph{information artifacts},
with adequate metadata that is indexed in standardized repositories (Van De Sompel, Payette, Erickson, Lagoze, \& Warner, 2004).
Data sharing is an example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
Scholars can archive information artifacts in repositories like \href{https://zenodo.org/}{Zenodo},
which was developed by \href{https://home.cern/}{CERN} under the European Union's \href{https://www.openaire.eu/}{OpenAIRE} program (European Organization For Nuclear Research \& OpenAIRE, 2013).
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for information artifacts - as does, for example, the \href{https://www.crossref.org/}{Crossref} association,
which is used by many academic publishers.
Finally, the DataCite Metadata Schema offers a standard way to document the nature of relationships between information artifacts (DataCite Metadata Working Group, 2024).
For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property \texttt{resourceType:\ dataset},
and cross-reference the published paper with \texttt{relationType:\ IsSupplementTo}.
Similarly, a FAIR theory object can be connected to a specific paper which might serve as the theory's documentation and canonical reference by using \texttt{relationType:\ IsDescribedBy}, while the reverse relationship, documented in the canonical reference paper, is \texttt{relationType:\ Describes}. Other types are useful for making relationships between multiple theory objects explicit: If an existing theory is made FAIR without substantial alterations,
the resulting FAIR theory metadata would cross-reference the existing theory as \texttt{relationType:\ IsDerivedFrom}.
If an existing theory is updated, \texttt{relationType:IsNewVersionOf} could be used to reference previous versions.
If a variation of an existing FAIR theory is created, cross-reference it with \texttt{relationType:\ IsVariantFormOf}.

@@

We can take inspiration from the field of computer science for well-established processes for iteratively improving information artifacts.
Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions from collaborators and allowing for experimentation and diverging development without losing information.
Git tracks line-by-line changes to text-based files,
and maintains a complete history of those changes.
It has long been argued that Git is particularly well-suited to academic work (Ram, 2013; Van Lissa et al., 2021).
Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration (Peikert, Van Lissa, \& Brandmaier, 2021; Van Lissa et al., 2021).
Git provides a useful framework for developing FAIR theory,
because it enables explicitly comparing versions of a file (or: theory),
incorporating changes by different authors,
and branching off into different directions (e.g., competing hypotheses) while retaining an explicit link to the common ancestor.
This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence (Van Lissa, 2023).
Note that archival of the version of record is \emph{not} a function of Git(Hub).
While theory development may take place on GitHub, the version of record should be archived in a FAIR-compliant archive like Zenodo, with appropriate metadata.

@@

The four Findability criteria are applicable to theory with only minor adjustments, see \href{https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv}{Supplemental Table S1}.
First, this requires assigning a globally unique and persistent identifier, such as a DOI, to each theory (F1).
Of the many services that provide DOIs for scientific information artifacts,
Zenodo and the Open Science Framework are commonly used in psychology.
Second, Findable theory is described with rich metadata (F2).
This includes citation metadata (e.g., referencing a scientific paper that documents the theory, or a psychometric paper that operationalizes specific constructs).
It might further include domain-specific metadata, such as a reference to a taxonomy of psychological constructs (Bosco, Uggerslev, \& Steel, 2017),
ontology (Guyon, Kop, Juhel, \& Falissard, 2018),
or catalog of psychological phenomena.
Metadata should also include identifiers for all the versions of the theory it describes (F3);
Zenodo handles this by default by providing an overarching DOI for an information artifact which subsumes the DOIs of that artifact's versions.

@@

There have been notable efforts to improve theories' Findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and posted many examples on their website \textless www.theorymaps.org\textgreater{} (Gray, 2017).
Similarly, \href{http://psychomodels.org/}{PsychoModels} seeks to inventorize theories and models in psychology (van Dongen \& Volz, 2025).
Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.
Both approaches can be complementary, however.
For example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
Conversely, the database engineering adage ``Lots of Copies Keeps Stuff Safe'' (LOCKSS) implies that it is fine to archive theories in multiple places,
although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places, which increases the risk of inconsistencies arising.

@@

Transparent scholarly communication about theory requires that theories are Accessible to all researchers and other stakeholders.
If theories are not Accessible, researchers cannot reuse and refine them.
Thus, Accessibility can accelerate cumulative knowledge acquisition.
Making theories Accessible also allows stakeholders (e.g., practitioners, policymakers, advocates) to inform themselves of the current scientific understanding of specific phenomena.
While isolated empirical findings can appear fragmented and contradictory (Dumas-Mallet, Smith, Boraud, \& Gonon, 2017),
theories offer a top-down, big-picture representation of the phenomena studied in a field.
In other words, theories are an important instrument in science communication.

@@

A second impediment to Accessibility is more indirect
and pertains to a theory's intelligibility to those with practical Access.
It has been proposed that good theories have the property of ``discursive survival {[}\ldots{]}, the ability to be understood'' (Guest, 2024).
At present, psychological theories are often ambiguous, rendering them difficult to understand (Frankenhuis et al., 2023).
Successful communication requires shared background knowledge between sender and receiver (Vogt et al., 2024).
Shared background knowledge can come from paradigms held by members of a scientific community (Kuhn, 2009),
from education, and from the available instrumentation for observation, measurement, and analysis - or it can be problematically absent.
Accessibility is improved by explicitly referring to sources of assumed backround knowledge, and by reducing unnecessary ambiguity.
At the same time, it is important to acknowledge the \emph{indeterminacy of translation}\footnote{Every communicative utterance (e.g., a statement in natural language, a mathematical formula, et cetera)
  has multiple alternative translations, with no objective means of choosing the correct one.},
which implies that it is not possible to remove \emph{all} ambiguity when communicating an idea (Quine, 1970).
This places a theoretical upper bound on theories' ability to be understood.

@@

In sum, authors should make good-faith efforts to make theories as Accessible as possible, in terms of both availability, intelligibility, and freedom from dependencies that cannot be resolved (including dependencies on the author, or manuscripts that can no longer be accessed with reasonable effort).
It is important to recognize that there is an upper bound on interpretability, which means that it is impossible to communicate a theory completely unambiguously.
Nevertheless, scholars should strive to reduce unnecessary ambiguity to the greatest possible extent.
It may benefit scientific discourse to normalize explicit ambiguity (these are things we don't know yet) and anticipate misunderstanding,
to invite others to fill in the blanks and motivate ever further explication of theory.
A theory's Accessibility is increased by
reducing dependencies on (implicit) background knowledge,
explication of assumptions,
formalization,
and explicit cross-references to relevant resources such as papers, ontologies, other related theories, measurement instruments, experimental designs (J. Lange, Freyer, Musfeld, Schönbrodt, \& Leising, 2025).

@@

Interdisciplinarity benefits from FAIR theory's Accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling (Haslbeck, Ryan, Robinaugh, Waldorp, \& Borsboom, 2021).
Meta-research benefits from the fact that FAIR theory enables studying the structure, content, and development of theories over time.
In terms of team science, FAIR theory facilitates collaboration by ensuring that all contributors have access to the same information and
clarifying any remaining areas of contention or misunderstanding (Van Lissa et al., 2024).
Version control provides a framework to resolve parallel developments from multiple collaborators in a non-destructive manner.
This facilitates collaboration across geographical boundaries,
and adversarial collaboration, where others strive to falsify a theory or identify its inconsistencies, and democratizes collaboration with as-of-yet unknown collaborators via platforms like GitHub, where researchers outside one's network can identify issues or suggest improvements to theories.

>>>>

  man, noextraspace,floatsintext]{apa7}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\makeatother
\keywords{fairtheory, meta science, theory formation, open science\newline\indent Word count: 9491}
\usepackage{lineno}

@@

This is a preprint paper, generated from Git Commit \# 5d7ad1a4. The first and senior author positions were fixed, and remaining author positions randomized using \texttt{set.seed(6999982);\ sample(c("Aaron",\ "Felix",\ "Noah",\ "Maximilian"))}.

The authors made the following contributions. Caspar J. Van Lissa: Conceptualization, Formal Analysis, Funding acquisition, Methodology, Project administration, Software, Supervision, Writing -- original draft, Writing -- review \& editing; Aaron Peikert: Conceptualization, Formal Analysis, Writing -- original draft, Writing -- review \& editing; Maximilian S. Ernst: Writing -- review \& editing, Investigation; Noah N.N. van Dongen: Writing -- review \& editing; Felix D. Schönbrodt: Conceptualization, Writing -- review \& editing; Andreas M. Brandmaier: Formal Analysis, Writing -- original draft, Writing -- review \& editing.

@@

\abstract{%
Innovations in open science and meta-science have focused on rigorous \emph{theory testing}, yet methods for specifying, sharing, and iteratively improving theories remain underdeveloped. To address these limitations, we introduce \emph{FAIR theory}: A standard for specifying theories as Findable, Accessible, Interoperable, and Reusable information artifacts. FAIR theories are Findable in well-established archives, Accessible in terms of availability and their ability to be understood, Interoperable for specific purposes, such as selecting control variables, and Reusable so that they can be iteratively improved through collaborative efforts. \phantomsection\label{abstractpackage}{This paper adapts the FAIR principles for theory, reflects on FAIR practices in contemporary psychology, introduces a workflow for FAIRifying theory, which is largely automated by the \texttt{theorytools} R-package}, and discusses FAIR theories' potential impact in terms of reducing research waste, enabling meta-research on theories' structure and development, and incorporating theory into reproducible research workflows -- from hypothesis generation to simulation studies. FAIR theory constitutes a protocol for archiving and communicating about theory, addressing a critical gap in open scholarly practices and supporting the renewed interest in theory development in psychology and beyond. FAIR theory builds on existing open science principles and infrastructures to provide a structured, cumulative framework for theory development, potentially increasing efficiency and potentially accelerating the pace of cumulative knowledge acquisition.
}

@@

The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable (M. D. Wilkinson et al., 2016) for both humans and computers\footnote{\phantomsection\label{faircap}{As the colloquial use of these terms differs in important ways from their definition according to the FAIR principles, we capitalize these terms when we refer to a specific FAIR principle.}}.
Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software (Lamprecht et al., 2019) and computational workflows (S. R. Wilkinson et al., 2024).
This paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
To this end, we introduce ``FAIR theory'':
a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
compliant with the FAIR principles.
\phantomsection\label{test2}{FAIR theory has the potential to improve the efficiency of scholarly communication and}
accelerate cumulative knowledge acquisition.
We focus on applications in psychology, but the principles are relevant across the social sciences and beyond.

@@

The so-called ``replication crisis'' has prompted extensive scientific reforms (Lavelle, 2021; Scheel, 2022).
Concern that the abundance of non-replicable findings
was caused by undisclosed flexibility in analyses led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
These various practices ensure transparent and repeated testing of hypotheses by committing to an analysis plan in advance.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence (Scheel, Schijen, et al., 2021).

Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so ambiguous that they can accommodate mutually inconsistent findings,
rendering them immune to falsification.
As an example, consider ``self-determination theory'' (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
Intrinsic motivation was initially defined as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures (Deci, 1971).
Over time, however, this definition expanded to include motivations driven by the fulfillment of basic psychological needs for autonomy, competence, and relatedness (Ryan \& Deci, 2000).
The implications of these shifting definitions becomes clear when deriving hypotheses about the type of motivation involved in changing an infant's dirty diaper.
Under the original definition, one would hypothesize that caregivers are not intrinsically motivated to change diapers, as this is hardly a joyous experience.
Under the expanded definition, one would hypothesize the opposite, as the act fulfills the need for relatedness.
The expanded definition thus enables SDT to absorb potentially falsifying evidence.

Scholars have raised concerns about the state of theory in psychology for nearly 50 years (Meehl, 1978; Robinaugh et al., 2021).
One main concern is that theories lack \emph{formalization} (Szollosi \& Donkin, 2021).
When theories are ambiguous and hence require either subjective interpretation or additional background knowledge,
it becomes difficult do derive precise predictions,
and therefore hard to falsify the theory.
A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
Despite these concerns, scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical processes of theory construction and improvement have been largely overlooked.
The present paper addresses this knowledge gap by applying, for the first time,
open science principles to psychological theory.
We apply introduce \emph{FAIR theory} to
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

In a progressive research program (Lakatos, 1971),
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature:
According to Kühberger et al. (2014), 89.6\% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, et al., 2021).
Only 15\% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis (McPhetres et al., 2021).
The remaining 85\% of deductive studies lacked an explicit connection between theory and hypothesis.
Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
in which case it would be straightforward and important to make these explicit (Fried, 2020; Norouzi et al., 2024).
Or, perhaps some hypotheses are reported as part of entrenched research practices (Gigerenzer et al., 2004),
but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected (Van Lissa et al., 2020).
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
Collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings (van Rooij \& Baggio, 2021).

Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely connected to, let alone derived from, theory.
Moreover, test results do not routinely contribute to the improvement or rejection of theories.
The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology ``lack the cumulative character of scientific knowledge. They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest'' (Meehl, 1978, p. 1).
The present paper seeks to make theory more tangible and practical by instantiating it as a digital ``object'' that scholars can access, reuse, and update in their daily workflows.

@@

Definitions of theory are abundant, and are the subject of extensive scholarly debate.
Given that a pluriformity of definitions are consistent with FAIR theory principles,
our suggested approach is not limited to any one particular definition.
Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic (Winther, 2021).
The syntactic view describes theories as ``sets of sentences in a given logical domain language'' (Winther, 2021, ch.~2),
acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
We recognize the syntactic view in Meehl's (1990) hierarchy of ever-more specific ``statements'' a theory might contain:
statements about the types of entities postulated (i.e., ontology),
statements about causal connections between those entities,
statements about the functional form of those connections,
and statements about their specific numerical values (Frankenhuis et al., 2023; Guest, 2024).
The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
It shifts the focus from theories as collections of sentences to mathematical models.
The term ``model'' is not uniquely defined within the literature;
models have been described as ``specific instantiations of theories, narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory'' (Fried, 2020, p. 336).
This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
The pragmatic view holds that there might not be one structure or definition of scientific theories, but instead, definitions differ across scientific domains.
It also argues that nonformal aspects (e.g.~commonly used analogies) and practices (e.g.~experimental designs) can be an important part of scientific theories.

@@

We propose FAIR theory as an instantiation of modular publishing (Kircz, 1998).
In most fields, the primary unit of scientific communication is the academic paper.
A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
Modular publishing is the practice of making each of these resources available as independent citable \emph{information artifacts},
with adequate metadata that is indexed in standardized repositories (Van De Sompel et al., 2004).
Data sharing is an example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
Scholars can archive information artifacts in repositories like \href{https://zenodo.org/}{Zenodo},
which was developed by \href{https://home.cern/}{CERN} under the European Union's \href{https://www.openaire.eu/}{OpenAIRE} program (European Organization For Nuclear Research \& OpenAIRE, 2013).
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for information artifacts - as does, for example, the \href{https://www.crossref.org/}{Crossref} association,
which is used by many academic publishers.
Finally, the DataCite Metadata Schema offers a standard way to document the nature of relationships between information artifacts (DataCite Metadata Working Group, 2024).
For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property \texttt{resourceType:\ dataset},
and cross-reference the published paper with \texttt{relationType:\ IsSupplementTo}.
Similarly, a FAIR theory object can be connected to a specific paper which might serve as the theory's documentation and canonical reference by using \texttt{relationType:\ IsDescribedBy}, while the reverse relationship, documented in the canonical reference paper, is \texttt{relationType:\ Describes}. Other types are useful for making relationships between multiple theory objects explicit: If an existing theory is made FAIR without substantial alterations,
the resulting FAIR theory metadata would cross-reference the existing theory as \texttt{relationType:\ IsDerivedFrom}.
If an existing theory is updated, \texttt{relationType:IsNewVersionOf} could be used to reference previous versions.
If a variation of an existing FAIR theory is created, cross-reference it with \texttt{relationType:\ IsVariantFormOf}.

@@

We can take inspiration from the field of computer science for well-established processes for iteratively improving information artifacts.
Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions from collaborators and allowing for experimentation and diverging development without losing information.
Git tracks line-by-line changes to text-based files,
and maintains a complete history of those changes.
It has long been argued that Git is particularly well-suited to academic work (Ram, 2013; Van Lissa et al., 2021).
Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration (Peikert et al., 2021; Van Lissa et al., 2021).
Git provides a useful framework for developing FAIR theory,
because it enables explicitly comparing versions of a file (or: theory),
incorporating changes by different authors,
and branching off into different directions (e.g., competing hypotheses) while retaining an explicit link to the common ancestor.
This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence (Van Lissa, 2023).
Note that archival of the version of record is \emph{not} a function of Git(Hub).
While theory development may take place on GitHub, the version of record should be archived in a FAIR-compliant archive like Zenodo, with appropriate metadata.

@@

The four Findability criteria are applicable to theory with only minor adjustments, see \href{https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv}{Supplemental Table S1}.
First, this requires assigning a globally unique and persistent identifier, such as a DOI, to each theory (F1).
Of the many services that provide DOIs for scientific information artifacts,
Zenodo and the Open Science Framework are commonly used in psychology.
Second, Findable theory is described with rich metadata (F2).
This includes citation metadata (e.g., referencing a scientific paper that documents the theory, or a psychometric paper that operationalizes specific constructs).
It might further include domain-specific metadata, such as a reference to a taxonomy of psychological constructs (Bosco et al., 2017),
ontology (Guyon et al., 2018),
or catalog of psychological phenomena.
Metadata should also include identifiers for all the versions of the theory it describes (F3);
Zenodo handles this by default by providing an overarching DOI for an information artifact which subsumes the DOIs of that artifact's versions.

@@

There have been notable efforts to improve theories' Findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and posted many examples on \href{https://www.theorymaps.org}{their website} (Gray, 2017).
Similarly, \href{http://psychomodels.org/}{PsychoModels} seeks to inventorize theories and models in psychology (van Dongen \& Volz, 2025).
Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.
Both approaches can be complementary, however.
For example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
Conversely, the database engineering adage ``Lots of Copies Keeps Stuff Safe'' (LOCKSS) implies that it is fine to archive theories in multiple places,
although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places, which increases the risk of inconsistencies arising.

@@

Transparent scholarly communication about theory requires that theories are Accessible to all researchers and other stakeholders.
If theories are not Accessible, researchers cannot reuse and refine them.
Thus, Accessibility can accelerate cumulative knowledge acquisition.
Making theories Accessible also allows stakeholders (e.g., practitioners, policymakers, advocates) to inform themselves of the current scientific understanding of specific phenomena.
While isolated empirical findings can appear fragmented and contradictory (Dumas-Mallet et al., 2017),
theories offer a top-down, big-picture representation of the phenomena studied in a field.
In other words, theories are an important instrument in science communication.

@@

A second impediment to Accessibility is more indirect
and pertains to a theory's intelligibility to those with practical Access.
It has been proposed that good theories have the property of ``discursive survival {[}\ldots{]}, the ability to be understood'' (Guest, 2024, p. 1).
At present, psychological theories are often ambiguous, rendering them difficult to understand (Frankenhuis et al., 2023).
Successful communication requires shared background knowledge between sender and receiver (Vogt et al., 2024).
Shared background knowledge can come from paradigms held by members of a scientific community (Kuhn, 2009),
from education, and from the available instrumentation for observation, measurement, and analysis - or it can be problematically absent.
Accessibility is improved by explicitly referring to sources of assumed backround knowledge, and by reducing unnecessary ambiguity.
At the same time, it is important to acknowledge the \emph{indeterminacy of translation}\footnote{Every communicative utterance (e.g., a statement in natural language, a mathematical formula, et cetera)
  has multiple alternative translations, with no objective means of choosing the correct one.},
which implies that it is not possible to remove \emph{all} ambiguity when communicating an idea (Quine, 1970).
This places a theoretical upper bound on theories' ability to be understood.

@@

In sum, authors should make good-faith efforts to make theories as Accessible as possible, in terms of both availability, intelligibility, and freedom from dependencies that cannot be resolved (including dependencies on the author, or manuscripts that can no longer be accessed with reasonable effort).
It is important to recognize that there is an upper bound on interpretability, which means that it is impossible to communicate a theory completely unambiguously.
Nevertheless, scholars should strive to reduce unnecessary ambiguity to the greatest possible extent.
It may benefit scientific discourse to normalize explicit ambiguity (these are things we don't know yet) and anticipate misunderstanding,
to invite others to fill in the blanks and motivate ever further explication of theory.
A theory's Accessibility is increased by
reducing dependencies on (implicit) background knowledge,
explication of assumptions,
formalization,
and explicit cross-references to relevant resources such as papers, ontologies, other related theories, measurement instruments, experimental designs (J. Lange et al., 2025).

@@

Interdisciplinarity benefits from FAIR theory's Accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling (Haslbeck et al., 2021).
Meta-research benefits from the fact that FAIR theory enables studying the structure, content, and development of theories over time.
In terms of team science, FAIR theory facilitates collaboration by ensuring that all contributors have access to the same information and
clarifying any remaining areas of contention or misunderstanding (Van Lissa et al., 2024).
Version control provides a framework to resolve parallel developments from multiple collaborators in a non-destructive manner.
This facilitates collaboration across geographical boundaries,
and adversarial collaboration, where others strive to falsify a theory or identify its inconsistencies, and democratizes collaboration with as-of-yet unknown collaborators via platforms like GitHub, where researchers outside one's network can identify issues or suggest improvements to theories.
